Extended Data

### 2022-05-22 doc/manuscript/Dow_Extended_Data.Rmd
reorder extended data figures to match order of presentation in text

##figures
EDF_map = "Extended Data Figure 1" 
EDF_growth_curves_all = "Extended Data Figure 2"  
EDF_climwin = "Extended Data Figure 3"
EDF_pheno_all = "Extended Data Figures 4-5"
EDF_pheno_Tmax_climwin = "Extended Data Figure 4"
EDF_pheno_Tmax_month = "Extended Data Figure 5"
EDF_tree_ring_figs= "Extended Data Figure 6-7"
EDF_tree_rings_TMIN = "Extended Data Figure 6"
EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"

@@

![**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`. Base map source is ggplot2.](tables_figures/chronology_map.png)

@@

![**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018. For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data. Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached. Solid lines represent DBH change attributable to stem growth.](tables_figures/growth_curves_all.png){ width=100% }

@@

![**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$. Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`). Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$). Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model. Critical temperature windows are listed in `r EDT_phenology_parameters`.](tables_figures/climwin_figure.png){ width=70% }

@@

![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

>>>>

##figures
EDF_growth_curves_all = "Extended Data Figure 1"  
EDF_climwin = "Extended Data Figure 2"
EDF_pheno_all = "Extended Data Figures 3-4"
EDF_pheno_Tmax_climwin = "Extended Data Figure 3"
EDF_pheno_Tmax_month = "Extended Data Figure 4"
EDF_map = "Extended Data Figure 5" 
EDF_tree_ring_figs= "Extended Data Figure 6-7"
EDF_tree_rings_TMIN = "Extended Data Figure 6"
EDF_tree_rings_TMAX_extended = "Extended Data Figure 7"

@@

![**Extended Data Figure 1 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018. For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data. Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached. Solid lines represent DBH change attributable to stem growth.](tables_figures/growth_curves_all.png){ width=100% }

@@

![**Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$. Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`). Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$). Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model. Critical temperature windows are listed in `r EDT_phenology_parameters`.](tables_figures/climwin_figure.png){ width=70% }

@@

![**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_AIC.png){ width=90% }

@@

![**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.](tables_figures/pheno_Tsensitivity_combo_patchwork_april.png){ width=90% }

@@

![**Extended Data Figure 5 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`. Base map source is ggplot2.](tables_figures/chronology_map.png)

### 2022-05-22 doc/manuscript/Dow_manuscript_Nature.Rmd
reorder extended data figures to match order of presentation in text

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height (*∆DBH*; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`)[@dorangeville_peak_2022;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`, `r EDF_pheno_all`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (_CTW_). 
_CTW_ was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The _CTW_ was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (_CTW_ *T~max~*) used as our independent variable.

@@

**Extended Data Figure 1 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`. Base map source is ggplot2.

**Extended Data Figure 2 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018. For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data. Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached. Solid lines represent DBH change attributable to stem growth.

**Extended Data Figure 3 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$. Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`). Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$). Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model. Critical temperature windows are listed in `r EDT_phenology_parameters`.

**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.

**Extended Data Figure 5 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.

>>>>

Using dendrometer band measurements taken throughout multiple growing seasons at the Smithsonian Conservation Biology Institute (SCBI; Virginia, USA; n = `r n_dendro_trees_SCBI` trees from 2011-2020) and Harvard Forest (Massachusetts, USA; n = `r n_dendro_trees_HF` trees from 1998-2003), we fit a logistic growth model[@mcmahon_general_2015] to determine the days of year (DOY) when 25, 50, and 75% annual diameter growth were achieved (*DOY~25~*, *DOY~50~*, *DOY~75~*), peak growing season length (*L~pgs~ = DOY~75~ - DOY~25~*), the maximum growth rate (*g~max~*), and total annual increment in diameter at breast height (*∆DBH*; Fig. `r schematic_summary`). 
This analysis was performed separately for ring- <!--(67 trees at SCBI and 138 trees at Harvard Forest)-->and diffuse porous <!--(52 trees at SCBI and 183 trees at Harvard Forest)-->species (`r EDT_species`), which differ in the seasonal timing of growth (`r  EDT_phenology_parameters`, `r EDF_growth_curves_all`)[@dorangeville_peak_2022;@tumajer_limitation_2022]. 
These stem-growth milestones were compared to canopy foliage phenology (measured at ecosystem level via remote sensing).

@@

Both canopy foliage phenology and the timing of stem growth shifted earlier as spring temperatures increased (Fig. `r DOY_timing`). 
We found a consistent effect of temperature (*T~max~* or *T~min~*) throughout the spring, but the strongest effects on stem growth timing were found using *T~max~* during a critical temperature window (_CTW_). 
_CTW_ was identified by assessing the correlation between weekly *T~max~* and *DOY~25~* for all combinations of consecutive weeks from January 1 to mean *DOY~25~* for each xylem architecture-site combination (`r EDF_climwin`).
The _CTW_ was defined as the weeks for which this correlation was strongest, and mean *T~max~* during this period (_CTW_ *T~max~*) used as our independent variable.

@@

**Extended Data Figure 1 | Phenological patterns of forest canopy greenness (top row) and stem growth of ring- and diffuse-porous trees, represented as both relative and cumulative fractions of total annual growth (middle and bottom rows, respectively), at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** In the top row, canopy greenness is characterized using the two band Enhanced Vegetation Index (EVI2), with each line representing a year between 2000 and 2018. For stem growth, each line represents the average growth over one year, as modeled based on a five-parameter logistic growth model to dendrometer band data. Dashed lines represent modeled DBH change which fell outside of the median DOY where predicted starting DBH (a) and predicted ending DBH (b) were reached. Solid lines represent DBH change attributable to stem growth.

**Extended Data Figure 2 | Landscapes of relationships between the day of year on which 25% of annual growth is achieved ($DOY_{25}$) and temperature in prior weeks for ring- and diffuse-porous trees at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** Shown are matrices of $\beta$ coefficients from first-order linear regressions between mean maximum temperature ($T_{max}$) and $DOY_{25}$. Window Open and Window Close indicate number of weeks prior to $DOY_{25}$ (listed in `r EDT_phenology_parameters`). Yellow shading indicates neutral relationships, while orange or red shading indicates that $DOY_{25}$ advances with increased $T_{max}$ over the given time window (negative $\beta$). Black circles indicate the critical temperature window selected based on minimization of $\Delta AICc$, the difference in Akaike Information Criterion corrected for small sample size relative to a null model. Critical temperature windows are listed in `r EDT_phenology_parameters`.

**Extended Data Figure 3 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) during the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`). Shown are relationships between mean $T_{max}$ over the _CTW_ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ all declined significantly with mean $T_{max}$ during their respective _CTW_. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.

**Extended Data Figure 4 | Response of stem growth timing and rates to mean maximum temperatures ($T_{max}$) for the month most closely corresponding to the spring critial temperature window (_CTW_) for ring- and diffuse-porous speces at the Smithsonian Conservation Biology Institute (SCBI) and Harvard Forest.** _CTW_ was defined as the period over which $T_{max}$ was most strongly correlated with the day of year on which 25% of annual growth was achieved ($DOY_{25}$; `r EDT_phenology_parameters`, `r EDF_climwin`), and the most closely corresponding month was determined as that with the greatest number of days within the _CTW_. Shown are relationships between montlhy $T_{max}$ and days of the year on which 25%, 50%, and 75% total stem growth were achieved ($DOY_{25}$, $DOY_{50}$, $DOY_{75}$, respectively; first row); the length of the peak growing season ($L_{pgs}$; second row); maximum growth rate ($g_{max}$; third row); and total seasonal radial stem growth ($\Delta DBH$; fourth row). Posterior predictions of each variable that did not include zero are represented with solid lines, while those that do include zero use dotted lines. The 95% credible intervals are represented by bands centered on the posterior mean for each year. For both species groups at both sites, $DOY_{25}$, $DOY_{50}$, and $DOY_{75}$ declined significantly with April $T_{max}$, with the exception of $DOY_{25}$ for ring-porous species at SCBI. Dots represent growth parameter values for individual tree-year combinations, which were derived by fitting a five-parameter logistic growth model to dendrometer band data.

**Extended Data Figure 5 | Map of sampling locations of tree-ring chronologies analyzed in this study.** Sites are colored by the xylem porosity type of species sampled: ring poroous (RP), diffuse porous (DP), or both. Sampling details are provided in `r chron_table`. Base map source is ggplot2.
